\thetitle

### 2020-07-02 writing/challenges.tex
polish challenges

\section{Challenges}

Creating a method for comparing blame assignment strategies poses several severe
challenges.

\cd{One point we haven's justified is why empirical. The justification
seems easy in the sense that we are looking at a real language, programs,
bugs rather than models.}

\cd{I like this idea of the competent programmer. I couldn't have thought
about homo economicus. Instead this was partly inspired by Gonen's work.}
\cd{There they make a hypothesis, try to test it in vivo using designed model organisms
with a well-specified first-order characteristic (e.g, a deleted gene).
Then they collect data by placing the organism in a number of scenarios 
and measuring aspects of its behavior and then they 
construct an operational model from their data. Due to the synthetic
nature of CS, for us the process goes in the opposite direction. In that
sense the competent programmer is a model organism that, as its
first-order characteristic, implements a direct consequence of the
semantics. So in a sense for me this has always been about semantics and
exploring their consequences.}

The most serious challenge concerns the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for research similar
to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a good number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm. As indicated in the introduction, in
the context of gradual typing a rationally acting programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion.

One derivative challenge to the first is the problem of simulating the
programmer's mistakes. Before a programmer can try to find mistakes, mistakes
have to be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in typed programs
with mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. This results into a large number of buggy programs whose
mixed-typed versions are suitable subjects for our experiment. However, mutation
analysis traditionally aims to inject bugs that are sophisticated enough to
challenge extensive test suites. As a consequence, the ill-typed mutants we care
about here are usually deemed \emph{incompetent} and mutation analysis
frameworks are fine-tuned to avoid them. To overcome this problem, we design a
set of mutators specific for type errors inspired by common errors when
programming in Typed Racket.  These type errors cover a wide range of Typed
Racket's type system and moreover cause reliably runtime type errors in
mixed-typed versions of mutants.

Another derivative challenge is the problem of mistakes in type specifications.
Gradual typing enables programmers to move an untyped code base into the typed
realm on an incremental basis. The retroactive addition of a specific type to a
component, even a library, may result from a misunderstanding of the code and
may thus be a mistake itself. Empirical evidence suggests that this scenario is
reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
hence must be addressed by any evaluation method. Similarly to the
previous challenge, we deal with this one by designing a set of mutators
that ``break'' type annotations but leave programs otherwise intact.

The second big challenge is the set of seemingly incomparable gradual
typing systems. Despite the fact that the natural, transient and optional
approaches are all well-understood as variants of the same minimal operational model
~\cite{gf-icfp-2018, gfd-oopsla-2019}, their actual implementations live on
different linguistic settings. Hence it is not possible to compare them directly 
on the same programs; code needs to be adjusted in often subtle ways to
transfer from one implementation to another. As a first step to overcome the problem, 
we have implemented a transient and optional variant of Typed Racket.  
However,
implementations on top of the same platform are not sufficient for an
apples-to-apples comparison. As we discuss in detail in
section~\ref{sec:landscape}, the three strategies differ in the kind of
blame information they produce. Natural blames one component while
transient blames a set of components and optional none. To account for these
differences, our competent programmer comes with \emp{modes} that represent 
the alternative ways programmers can interpret blame information. For instance, in
transient, one mode  has the competent programmer pick the oldest element of a blame set 
as the next component to add types. After all, this corresponds to the
earliest point in the program's evaluation that can discover the bug. 
Another mode though, does not distinguish between the elements of the
blame set and instead it requires that the competent programmer types 
all blamed components at once. In order to compare different strategies
across their different modes, we normalize the steps the competent
programmer takes to find the bug by the effort of each step, that is the
number of component it has to type. Finally, besides differences in how
and what they blame, the three gradual typing approaches differ in the type checks they perform. 
To untangle the two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
ignores blame and instead uses information from dynamic type error messages to
select the next component to type. Hence it separates the effect of blame 
from that of the type checks. The \emph{control} mode randomly
selects which part of the program to migrate and serves as the null hypothesis
control for the overall method.

>>>>

\section{Why is it hard to evaluate the value of blame?}

Creating a method for comparing blame assignment strategies poses several
challenges.

% \cd{One point we haven's justified is why empirical. The justification
% seems easy in the sense that we are looking at a real language, programs,
% bugs rather than models.}
%% MF: I edited the text to include this.

% \cd{I like this idea of the competent programmer. I couldn't have thought
% about homo economicus. Instead this was partly inspired by Gonen's work.}
% \cd{There they make a hypothesis, try to test it in vivo using designed model organisms
% with a well-specified first-order characteristic (e.g, a deleted gene).
% Then they collect data by placing the organism in a number of scenarios 
% and measuring aspects of its behavior and then they 
% construct an operational model from their data. Due to the synthetic
% nature of CS, for us the process goes in the opposite direction. In that
% sense the competent programmer is a model organism that, as its
% first-order characteristic, implements a direct consequence of the
% semantics. So in a sense for me this has always been about semantics and
% exploring their consequences.}

%% MF says: yes, I figured you were inspired by biology but in a
% synthetic world, like econ, you need also a rationale for why this
% thing is worthwhile measuring .. and so they came up with homo
% econ.

The most serious challenge concerns the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs is the equivalent of
collecting empirical data in a truly human-focused discipline.

One derivative challenge of the first one is the problem of simulating the
programmer's mistakes or, more precisely, the origin of impedance
mismatches. Before a programmer can try to find mistakes, mistakes have to be
made. Technically the problem is the lack of a sufficiently large available
corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, Following~\citet{lksfd-popl-2020} again, the solution
is to use mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis} to create mistakes in a synthetic manner. This process yields a
large number of buggy programs whose mixed-typed versions are suitable subjects
for our experiment.  The mutation analysis of Lazarek et al, is, however,
completely useless for creating impedance mismatches in a gradually typed
context.

While mutation analysis traditionally aims to inject bugs that challenge
extensive test suites, the ones yielding ill-typed mutants are usually deemed
\emph{incompetent}. As a matter of fact, mutation analysis frameworks are
fine-tuned to avoid them---yet it is precisely those that are needed for
evaluating type impedance mismatches. The ones used here are newly designed,
inspired by the authors extensive experience with programming in Typed Racket.
These type errors cover a wide range of Typed Racket's type system and moreover
reliably cause runtime exceptions in mixed-typed versions of mutants.

Another derivative challenge of the first one is the problem of mistakes in type
annotations themselves.  Gradual typing enables programmers to move an untyped
code base into the typed realm on an incremental basis. The retroactive addition
of a specific type to a component, even a library, may result from a
misunderstanding of the code and may thus be a mistake itself. Empirical
evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
method. Conversely, the method not only demands mutators for code that cause
good impedance mismatches, it also calls for a set of mutators that ``break''
type annotations without breaking the code itself.

The second big challenge is due to implementation disparity of gradual type
systems. While the various approaches are well understood in theory, they come
with vastly different type system, accommodate seemingly incompatible language
features, and apply to radically different coding styles in the untyped world.
To eliminate these differences as variables of a comparison, the application of
an evaluation method must use the same test programs and the same type system.
The case study presented here builds on Greenman and Felleisen's transient
variant for functional Typed Racket~\cite{gf-icfp-2018}. the revised transient
variant accommodates many more linguistic features and calls library code in a
way that understand the transient semantics.

Beyond the implementation issue, a comparison also faces the problem that the
three variants---the natural semantics, the transient semantics, and the erasure
semantics---come with three radically different blame assignment systems.  For
example, the natural semantics assigns blame to one component while transient
assigns blame to a set of components. The erasure semantics does not blame a
component per se, but it comes with an exception location and the usual stack
traces.

Making these strategies comparable is {\em not\/} an issue of language
implementation but one about how the rational programmer behaves with regard to
these messages. Specifically, the rational programmer comes with \emph{modes}
that represent the alternative ways in which a programmer may interpret blame
information. For instance, in transient, one mode has the competent programmer
pick the oldest element of a blame set as the next component to add types. After
all, this corresponds to the earliest point in the program's evaluation that can
discover the bug.  Another mode does not distinguish between the elements of the
blame set and requires that the rational programmer types all blamed components
at once.  Once information about the search steps in each mode is available, it
is normalized. In particular, it is not just the steps that are counted but the

Moreover, the three gradual typing systems differ in the type checks they
perform.  An extra mode aims to untangle the two issues. This \emph{vanilla} mode
ignores blame and instead uses information from dynamic type error messages to
select the next component to type. Hence it separates the effect of blame from
that of the type checks.

In sum, the peculiarities of the gradual typing landscape demand a careful
design of both the evaluation method and the application of the method to
specific gradual typing systems. Otherwise measuring the rational programmer
will yield potentially random results.
